Data-Driven Modelling \& \\ Forecasting of Chaotic Systems

### 2022-01-19 projectReport/chapter2.tex
Rereading the project. On 5.4

\begin{Definition}
  [\bf Attractor]\label{Dfn_Attractor}\rm
  Let $T: U \to U$, where $U$  is a metric space with metric $d$. A compact subset $A \subset U$ is said to be an attractor if it satisfies the three conditions: 
  \vspace{-8mm}
  \begin{enumerate}
	\item $A$ is invariant. 
	\item $A$ is asymptotically stable, i.e., for every $\epsilon > 0$ and for all $u$ so that $d(u,A) < \epsilon$, we have $d(T^nu,A){\to}{0}$ as $n\to \infty$. 
	\item $A$ has Lyapunov stability, i.e., for every $\epsilon > 0$  there exists a $\delta(\epsilon) > 0$ so that $d(u,A) < \delta$ implies $d(T^nu,A) < \epsilon$ for all $n\ge 0$.  
\end{enumerate}
\end{Definition}

@@

To this end, we make 3 remarks:
\vspace{-5mm}
\begin{enumerate}
  \item The $\delta>0$ in the definition of SDIC is independent of $u$. 
  \item In every neighborhood of $u$, we may not necessarily find all points $v$ in the neighborhood distinct from $u$ that would separate from the forward iterates of $u$. 
  \item  $N$ depends upon $u$ and $v$ chosen, and their iterates may not separate forever (i.e., for all $n>N$) and we allow their iterates to get arbitrarily close in the future. 
\end{enumerate}

@@

\begin{Example}\label{ex_LM} \rm
  The logistic map (LM), a recursion relation of the form $x_{n+1}=rx_n(1-x_n)$ where $x_n\in[0,1]$ and $r\leq{4}$, is a classic example used to illustrate the chaotic behaviour that a system may exhibit. When $r=4$, the one-dimensional recurrence relation $x_{n+1}=4x_n(1-x_n)$ can be used as the kindergarten's model to exhibit the presence of SDIC. %\ednote{B:edited and removed unnecessary citation}
  In Graph~\ref{fig:log_sdic} below, we plot the first 50 iterates of the LM for $r=4$ and two slightly different initial values $x_0$. 
  The values of $x_n$ are plotted agains time $n$  where they have initial values $x_0=0.2$ (blue) and $x_0=0.20001$ (red). Initially the two trajectories overlap, but they diverge completely at $n=12$ whereafter they follow distinct paths.

@@

We may now define the notion of Chaos as formulated by Devaney\cite{devaney2018introduction}.
\begin{Definition}
  [\bf {Devaney's Chaos}]\label{Dfn_ChaosDec}\rm
	A dynamical system $T: U \to U$ is said to exhibit chaos in the sense of Devaney if it satisfies the three properties:
	\vspace{-5mm}
  \begin{enumerate}
		\item $T$ has SDIC.
		\item $T$ is topologically transitive.
		\item The set of periodic points of $T$ are dense in $U$. 
	\end{enumerate}
\end{Definition}

@@

    In general, the graph of  $T^k$ contains $2^{k-1}$ tents.
    Given any point $u$ and an open neighborhood $W \subset[0,1]$ of $u$, we can find $k\in\mathbb{Z}$ large enough so that $T^k(W) = [0,1]$ because we can accommodate a tent whose base is contained in $W$. % \ednote{B: This last phrase doesn't make sense to me. M: The number of tents grow with $k$ and hence this is obvious; I explained in one of our meetings. Does it make sense?} 
    This implies the existence of two points $w_1$ and $w_2$ in $W$ so that $d(T^kw_1,T^kw_2)= 1$, and by the triangle inequality  at least one of the inequalities $d(T^ku,T^kw_1)> \delta$ or $d(T^ku,T^kw_2)> \delta$ hold when $\delta\in (0,\frac{1}{2})$.  So $T$ has sensitive dependence on initial conditions.

@@

We now turn our attention to the subject of conjugacy that describes when two dynamical systems are equivalent dynamically.

To show topological similarity (or sameness) between two metric or topological spaces, one needs to establish a homeomorphism between the two spaces. 
However, in the study of dynamical systems defined on two spaces, establishing a homeomorphism does not indicate that the systems are dynamically related in any way.
For instance, the maps $Tu=u^2$ and $Tu=1-|2u-1|$ defined on $[0,1]$ have totally different behaviour. 
Therefore, to find dynamically similar systems, one must establish a dynamical equivalence that we illustrate in a simple commutativity diagram below~\ref{eqn_conjugacy}.

% Giving errors
\begin{equation}  \label{eqn_conjugacy}
%    \[ 
    \psset{arrows=->, arrowinset=0.25, linewidth=0.6pt, nodesep=3pt, labelsep=2pt, rowsep=0.7cm, colsep = 1.1cm, shortput =tablr}
 \everypsbox{\scriptstyle}
 \begin{psmatrix}
U & U\\%
V & V.
 %%%
%  \ncline{1,1}{1,2}^{T} 
%  \ncline{1,1}{2,1} <{\phi}
%   \ncline{2,1}{2,2}^{S}
%  \ncline{1,2}{2,2} > {\phi}
 \end{psmatrix}
% \]
\end{equation}

%Composite functions
To understand the diagram, we note that if we travel `right and then down', the diagram instructs us to use $T$ (top arrow right) first, followed by $\phi$ (right arrow downwards). Consequently, the pathway amounts to finding $\phi(T(u))$. If we proceed `down and then right', the diagram instructs us to apply $\phi$ (left arrow downwards) first, and then apply $S$ (bottom arrow right). This then amounts to finding  $S(\phi(u))$. When the relationships in this diagram hold, we say $\phi(T(u))= S(\phi(u))$, and we formally denote it as $\phi \circ T=S\circ \phi$.

\begin{Definition}
  [\bf {Conjugacy and Semi-Conjugacy}]\label{Dfn_Conjugate}\rm
   Consider the dynamical systems $T:U\to{U}$, $S:V\to{V}$ and suppose the relationship $\phi \circ T=S\circ \phi$ holds. If $\phi:U\to{V}$ is a homeomorphism, then $S$ is said to be conjugate to $T$ and $\phi$ is called a conjugacy. If we relax the criterion on $\phi$ and merely require $\phi$ to be continuous and surjective where $\phi:U\to{V}$, then $\phi$ is a semi-conjugacy between $T$ and $S$ where T has domain $U$ and S has domain $V$; S is said to be semi-conjugate to T. 
\end{Definition}

\ednote{B: Does defining explicitly the domain and codomain of $\phi$ make it clear enough?  M: Saying $\phi$ is then a semi-conjugacy between $T$ and $S$ could be ambiguous as it does not specify which system is on the top in the commutativity. So please edit accordingly.}
When $S$ is conjugate to $T$, the dynamics of the two systems are in some way `dynamically equivalent'. Specifically, they are in one-to-one correspondence with one another. However when $S$ is semi-conjugate to $T$ with $\phi:U\to{V}$ a many-to-one mapping, the dynamics on $V$ provide merely a coarse-grained description of the dynamics on $U$~\cite{de2013elements}. When $S$ is semi-conjugate to $T$, it is also common to call $S$ a \emph{factor of $T$}, or conversely that $T$ is an \emph{extension of $S$}. In essence, an extension (see,~\cite{de2013elements}) is a more extensive system capturing all of the essential dynamics of its factor.

It is a very hard or nearly an impossible task to establish the existence of a conjugacy or a semi-conjugacy $\phi$ between two systems~\cite{devaney2018introduction}. However, one can verify that a function $\phi$ satisfies the commutativity diagram~\ref{eqn_conjugacy}.

\begin{Example}\rm
  For example $\phi(x)=\sin{\big(\frac{\pi}{2}x\big)}^2$ is a conjugacy between two systems $Tu=1-|2u-1|$ and $Sv=4v(1-v)$  defined on $U=[0,1]$.

  Indeed, $\phi:[0,1]\to[0,1]$ is a homeomorphism and by employing simple trigonometric identities, it is easily proved that $\phi$ satisfies~\ref{eqn_conjugacy}
  \[
    \begin{aligned}[t]
      (\phi\circ{T})u 
                &=\sin^{2}\Big(\frac{\pi}{2} - \frac{\pi}{2}|2u-1|\Big) = \sin^{2}(\pi{u})={\Big(2\sin(\frac{\pi}{2})\cos(\frac{\pi}{2})\Big)}^{2}=4\sin^{2}(\frac{\pi}{2}u)\Big(1-\sin^{2}\frac{\pi}{2}u\Big) \\
                                    &=({S}\circ\phi)u.
    \end{aligned}
\]
And it follows that $\phi$ is a conjugacy between the two systems.
\end{Example}

By establishing that two systems are conjugate (semi-conjugate) to one another, we have also shown that one may choose to work with one system instead of the second and still be guaranteed to obtain information on the latter. It is instrumental during the process of forecasting the future evolution of dynamical systems.

>>>>

\begin{Definition}
  [\bf Attractor]\label{Dfn_Attractor}\rm
  Let $T: U \to U$, where $U$  is a metric space with metric $d$. A compact subset $A \subset U$ is said to be an attractor if it satisfies the three conditions.
  \vspace{-8mm}
  \begin{enumerate}
	\item $A$ is invariant. 
	\item $A$ is asymptotically stable, i.e., for every $\epsilon > 0$ and for all $u$ so that $d(u,A) < \epsilon$, we have $d(T^nu,A){\to}{0}$ as $n\to \infty$. 
	\item $A$ has Lyapunov stability, i.e., for every $\epsilon > 0$  there exists a $\delta(\epsilon) > 0$ so that $d(u,A) < \delta$ implies $d(T^nu,A) < \epsilon$ for all $n\ge 0$.  
\end{enumerate}
\end{Definition}

@@

To this end, we make 3 remarks.
\vspace{-5mm}
\begin{enumerate}
  \item The $\delta>0$ in the definition of SDIC is independent of $u$. 
  \item In every neighborhood of $u$, we may not necessarily find all points $v$ in the neighborhood distinct from $u$ that would separate from the forward iterates of $u$. 
  \item  $N$ depends upon $u$ and $v$ chosen, and their iterates may not separate forever (i.e., for all $n>N$) and we allow their iterates to get arbitrarily close in the future. 
\end{enumerate}

@@

\begin{Example}\label{ex_LM} \rm
  The logistic map (LM), a recursion relation of the form $x_{n+1}=rx_n(1-x_n)$ where $x_n\in[0,1]$ and $r\leq{4}$, is a classic example used to illustrate the chaotic behaviour that a system may exhibit. When $r=4$, the one-dimensional recurrence relation $x_{n+1}=4x_n(1-x_n)$ can be used as the kindergarten's model to exhibit the presence of SDIC.
  In Graph~\ref{fig:log_sdic} below, we plot the first 50 iterates of the LM for $r=4$ and two slightly different initial values $x_0$. 
  The values of $x_n$ are plotted agains time $n$  where they have initial values $x_0=0.2$ (blue) and $x_0=0.20001$ (red). Initially the two trajectories overlap, but they diverge completely at $n=12$ whereafter they follow distinct paths.

@@

We may now define the notion of Chaos as formulated by Devaney\cite{devaney2018introduction}.
\begin{Definition}
  [\bf {Devaney's Chaos}]\label{Dfn_ChaosDec}\rm
	A dynamical system $T: U \to U$ is said to exhibit chaos in the sense of Devaney if it satisfies the three properties.
	\vspace{-5mm}
  \begin{enumerate}
		\item $T$ has SDIC.
		\item $T$ is topologically transitive.
		\item The set of periodic points of $T$ are dense in $U$. 
	\end{enumerate}
\end{Definition}

@@

    In general, the graph of  $T^k$ contains $2^{k-1}$ tents.
    Given any point $u$ and an open neighborhood $W \subset[0,1]$ of $u$, we can find $k\in\mathbb{Z}$ large enough so that $T^k(W) = [0,1]$ because we can accommodate a tent whose base is contained in $W$.
    This implies the existence of two points $w_1$ and $w_2$ in $W$ so that $d(T^kw_1,T^kw_2)= 1$, and by the triangle inequality  at least one of the inequalities $d(T^ku,T^kw_1)> \delta$ or $d(T^ku,T^kw_2)> \delta$ hold when $\delta\in (0,\frac{1}{2})$.  So $T$ has sensitive dependence on initial conditions.

@@

We now turn our attention to the subject of conjugacy to describe when two dynamical systems are dynamically equivalent.

To show topological similarity (or sameness) between two metric or topological spaces, one needs to establish a homeomorphism between these two spaces. 
However, in the study of dynamical systems, establishing a homeomorphism does not indicate that the systems are dynamically related in any way.
For instance, the maps $Tu=u^2$ and $Tu=1-|2u-1|$ defined on $[0,1]$ have totally different behaviour. 
Therefore, to find dynamically similar systems, one must establish a dynamical equivalence as illustrated in the simple commutativity diagram below~\ref{eqn_conjugacy}.

\begin{equation}\label{eqn_conjugacy}
  \begin{tikzcd}
    U && U \\
    V && V \\
    \arrow["T", from=1-1, to=1-3]
    \arrow["\phi", from=1-3, to=2-3]
    \arrow["\phi"', from=1-1, to=2-1]
    \arrow["S"', from=2-1, to=2-3]
  \end{tikzcd}
  \end{equation}

%Composite functions
To understand this diagram, we note that if we travel `right and then down', the diagram instructs us to apply $T$ (top arrow right) first, followed by $\phi$ (right arrow downwards). Consequently, the pathway amounts to finding $\phi(T(u))$. If we proceed `down and then right', the diagram instructs us to apply $\phi$ (left arrow downwards) first, and then apply $S$ (bottom arrow right). This then amounts to finding  $S(\phi(u))$. When the relationships in this diagram hold, we say $\phi(T(u))= S(\phi(u))$, and we formally denote it as $\phi \circ T=S\circ \phi$.

\begin{Definition}
  [\bf {Conjugacy and Semi-Conjugacy}]\label{Dfn_Conjugate}\rm
   Consider the dynamical systems $T:U\to{U}$, $S:V\to{V}$ and suppose the relationship $\phi \circ T=S\circ \phi$ holds. If $\phi:U\to{V}$ is a homeomorphism, then $S$ is said to be conjugate to $T$ and $\phi$ is called a conjugacy. If we relax the criterion on $\phi$ and merely require $\phi$ to be continuous and surjective where $\phi:U\to{V}$, then $\phi$ is a semi-conjugacy between $T$ and $S$, where T has domain $U$ and S has domain $V$; S is said to be semi-conjugate to T. 
\end{Definition}

% \ednote{B: Does defining explicitly the domain and codomain of $\phi$ make it clear enough?  M: Saying $\phi$ is then a semi-conjugacy between $T$ and $S$ could be ambiguous as it does not specify which system is on the top in the commutativity. So please edit accordingly.}
When $S$ is conjugate to $T$, the dynamics of the two systems are in some way `dynamically equivalent'. Specifically, they are in one-to-one correspondence with one another. However when $S$ is semi-conjugate to $T$ with $\phi:U\to{V}$ a many-to-one mapping, the dynamics on $V$ provide merely a coarse-grained description of the dynamics on $U$~\cite{de2013elements}. When $S$ is semi-conjugate to $T$, it is common to call $S$ a \emph{factor of $T$}, or conversely that $T$ is an \emph{extension of $S$}. In essence, an extension (see,~\cite{de2013elements}) is a more extensive system capturing all of the essential dynamics of its factor.

It is a very hard or nearly an impossible task to establish the existence of a conjugacy or a semi-conjugacy $\phi$ between two systems~\cite{devaney2018introduction}. However, one can more easily verify that a function $\phi$ satisfies the commutativity diagram~\ref{eqn_conjugacy}.

\begin{Example}\rm
  $\phi(x)=\sin{\big(\frac{\pi}{2}x\big)}^2$ is a conjugacy between two systems $Tu=1-|2u-1|$ and $Sv=4v(1-v)$  defined on $U=[0,1]$.

  Indeed, $\phi:[0,1]\to[0,1]$ is a homeomorphism and by employing simple trigonometric identities, it is easily proved that $\phi$ satisfies~\ref{eqn_conjugacy}
  \[
    \begin{aligned}[t]
      (\phi\circ{T})u 
                &=\sin^{2}\Big(\frac{\pi}{2} - \frac{\pi}{2}|2u-1|\Big) = \sin^{2}(\pi{u})={\Big(2\sin(\frac{\pi}{2})\cos(\frac{\pi}{2})\Big)}^{2}=4\sin^{2}(\frac{\pi}{2}u)\Big(1-\sin^{2}\frac{\pi}{2}u\Big) \\
                                    &=({S}\circ\phi)u.
    \end{aligned}
\]
So $\phi$ is a conjugacy between the two systems.
\end{Example}

By establishing that two systems are conjugate (semi-conjugate) to one another, we have simultaneously shown that one may choose to work with one system instead of the second and still be guaranteed information pertaining to the latter. This is instrumental during the process of forecasting the future evolution of dynamical systems.

### 2022-01-19 projectReport/chapter3.tex
Rereading the project. On 5.4

In this chapter, we acquaint ourselves with the question of forecasting dynamical systems with unknown underlying structures, state and discuss the shortcomings of the Takens delay embedding theorem and discuss various issues faced while forecasting.

Consider a relatively simple learning problem: 
Given the sequence $(u_0, u_1, \ldots, u_m)$ for $m\in\mathbb{Z}$, a finite segment of an orbit of the map $T$ where $T$ is defined by the update equation $u_{n+1} = Tu_n$, forecast the values $u_{m+1}, u_{m+1}$ where the map $T$ is unknown, given that $u_m$.

A practical example of this would be the subsequent scenario: given the time-sequential coordinates of an object moving in space, predict the future positions of that object. We are very rarely, if at all, presented with a problem where the entire state information is available to us in the form $u_n$ at some timestep $n\in\mathbb{N}$. Consider then a more complicated learning problem:

Suppose we only have the observations $\theta(u_0), \theta(u_1), \ldots, \theta(u_m)$ of the true system states $u$ in an unknown dynamical system $(U,T)$ and we wish to predict the values $\theta(u_{m+1}), \theta(u_{m+2})$ and so forth.

First we establish the method of Takens delay embedding. 
In this method, one considers a discrete-time dynamical system defined on a `nice' space (a smooth manifold)\ednote{B: Where can I find a citation for this?} that can be obtained as time-$K$ map of a continuous time-dynamical system, concepts which are formalised below.
We make observations from such a system, i.e., we observe the evolution of $\theta(w_0)$ (where $w_0$ represents the initial state) by examining  a finite set of values of $u_n :=  \theta(w_n) = \theta(Tw_{n-1})$. The sequence $\{u_n\}$ represents a scalar time-series and intuitively $\theta$ represents a probe inserted into a bigger system which is itself only measuring/extracting a small part of the greater system state at time $t=n$. 
Consider, for example, a thermometer erected to measure the ambient temperature in a local village. This measurement function, the thermometer, is capturing only a single aspect, the temperature, of a much grander dynamical system entailing the current weather of the surrounding area. Even more than that though- it is measuring a miniscule part of the global weather system.

@@

However, the actual state $u$ of a system is seldom, if ever, fully known. The most one can do is to insert probes into a system to obtain partial information through the measurements taken. Moreover, the process of taking a measurement itself introduces two additional aspects which complicate the problem: \ednote{B: Perhaps this edit makes more sense? M: Maybe replace difficulty by saying that we need to consider two aspects. }
\vspace{-8mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item A series of measurements over a specific time interval is inherently a discretisation procedure of the underlying continuous-time dynamical system. (Hence why we restricted our attention to discrete-time systems in the preceding section)
  \item A probe will never be entirely accurate, and so the act of measurement introduces a certain measure of numerical noise/inaccuracy.
\end{enumerate}

From here we construct a multidimensional observable using the method of stacking previous observations, i.e., we create delay-coordinate map defined by
$\Phi_{k,\theta}(w) := (\theta(T^{-k}w)\ldots,\theta(T^{-1}w),\theta(w))$.  
The concept of an observable is understood in the sense of an observable as originally introduced by~\cite{takens1981detecting, genericObservableAeyels} to refer to a probe or measurement function inserted into the system. 
Takens' theorem, in essence, refers to the fact that for a sufficiently large $k$, we can define a dynamical system on the space $\mathbb{R}^{k+1}$ with states $\Phi_{k,\theta}(w)$, $\Phi_{k,\theta}(Tw)$, $\Phi_{k,\theta}(T^2w)$, etc. This dynamical system is topologically conjugate to the unknown underlying system $(W,T)$. We recall the Takens delay embedding theorem next.

@@

We define first the concepts of homeomorphism and embedding:
\begin{Definition}
  {\bf(Homeomorphism)}\label{Dfn_homeo}\rm
  A homeomorphism is a function $f:Z\rightarrow Y$ between two topological spaces $Z$ and $Y$ that is continuous, bijective and has a continuous inverse. 
\end{Definition}

@@

Takens Theorem states a result establishing a relationship between the observed and underlying dynamical systems by showing that the concatenation of a sufficiently large number of previous observations into a vector will, under certain conditions, generate a map between the vectors from the respective systems.  We formulate the theorem from~\cite{takens1981detecting} after establishing preliminary definitions.

\ednote{It was suggested that I define these two concepts before stating the theorem.}
\begin{Definition}
  [\bf{Manifold, Compact Manifold}]\label{Dfn_compactManif}\rm
  A topological space is called a manifold if every point in the space has a neighbourhood homeomorphic to $\mathbb{R}^n$. Succinctly, a manifold is a locally Euclidean topological space.
  If a manifold is compact as a space topologically, then it is termed a compact manifold
\end{Definition}

@@

\begin{Theorem} 
	[\bf Takens Embedding Theorem (adopted from~\cite{takens1981detecting})]\label{Thm_Takens}
         Let $W$ be a compact manifold of dimension $m$, and $d\ge m$ so that $2d$ is an integer. It is a 
            generic property for the pair $(T, \theta)$,  where $T:W \to W$ is
            a smooth diffeomorphism, and $\theta:W \to \mathbb{R}$ a smooth function, the map $\Phi_{2d,\theta}:W \to \mathbb{R}^{2d+1}$ defined on $W$ by 
            $\Phi_{2d,\theta}(w) := (\theta(T^{-2d}w)\ldots,\theta(T^{-1}w),\theta(w))$
            is a diffeomorphic embedding. Consequently, there exists a map $F_\theta: \Phi_{2d,\theta}(W) \to \Phi_{2d,\theta}(W)$ defined by \[F_\theta: (\theta(T^{-2d}w),\ldots,\theta(T^{-1}w),\theta(w)) \mapsto 
            (\theta(T^{-2d+1}w),\ldots,\theta(w),\theta(Tw)) \]
           so that $(W,T)$ is topologically conjugate to 
            $(\Phi_{2d,\theta}(W), F_\theta)$.    
\end{Theorem}

By generic, we mean a residual set on a certain topology on appropriate spaces of functions (that we do not describe here) \ednote{B: Where can I find a citation for this?}. 
In our scenario (and this remains important throughout) the input space $U$ is considered the attractor.~\label{attractor_U}

@@

\begin{figure}[ht]
  \includegraphics[scale=0.25]{Graphs/_takensmap.eps}
  \centering
  \captionof{figure}{Schematic of Takens Embedding Theorem's usage and impact of iterates slipping from the attractor when the approximation $\tilde{F}_\theta$ of $F_\theta$ is learnt.}
 \label{fig:takensmap}
\end{figure}

@@

\section{Practical Limitations to Takens' Theorem}\label{sect_takenslimits}
Takens' Embedding Theorem is a powerful result, and it provides compelling reason to believe that one could conceivably reconstruct a system conjugate accurately to the underlying system. 
Nonetheless, it does present some severe practical limitations:
\vspace{-5mm}
\begin{enumerate}
\item Even supposing that we could indeed find $F_\theta$, our \emph{approximation} of  $F_\theta$ is a map from a larger set $\mathbb{R}^{2d+1}$ containing the embedded attractor. There are, however, no theoretical guarantees that $F_\theta$ will retain  $V=\Phi_{2d,\theta}(W)$  as an attractor although $W$ could itself be an attractor.
\item Takens theorem is stated only for noiseless observations. Due to noise $\epsilon_n$ the delay vector $\Phi_{2d,\theta}(w_n) + \epsilon_n$  may lie outside $V$.Furthermore, due to the chaotic nature of the underlying system (i.e. the fact that it has SIDC), the evolution of $\Phi_{2d,\theta}(w_n) + \epsilon_n$ under the map $F_\theta$ could move out of $V$ completely. This problem can be overcome  by using a driven dynamical system with some properties, and we discuss this in Chapter\ref{ch3}. 
\end{enumerate}

@@

Though not a uniquely related to the Takens embedding theorem, we do opt to make mention of another requirement for our work. 
The methods described here are applicable for data originating from a surjective map. 
This follows from the fact that when $T$ maps the space $U$ into a proper subset $B$ of$ $U, there exists some point in $U - A$ that does not have a preimage and we conclude that $T$ is not surjective. 
An example of this would be the case when a system displays energy-loss. Consider, for instance, consider the dying oscillations for a damped pendulum. (An account of greater depth  is provided in~\ref{ch5})

This requirement is not overly restrictive, as a great number of chaotic systems do have a surjective map. \textbf{(Cite)} \ednote{B: I remember you saying this in one of our sessions, but don't have a source. Could you point me to a resource?}

>>>>

In this chapter, we acquaint ourselves with the question of forecasting dynamical systems that have unknown underlying structures, state and discuss the shortcomings of the Takens delay embedding theorem, and discuss various issues faced while forecasting.

Consider a relatively simple learning problem:
Given the sequence $(u_0, u_1, \ldots, u_m)$ for $m\in\mathbb{Z}$, a finite segment of an orbit of the map $T$ where $T$ is defined by the update equation $u_{n+1} = Tu_n$, forecast the values $u_{m+1}, u_{m+2}, \ldots$ where the map $T$ is unknown.

A practical example of this would be to predict the future positions of an object moving in space given the time-sequential coordinates of that object. 
We are rarely, if ever, presented with a problem where the entire state information is available in the form $u_n$ at some timestep $n\in\mathbb{N}$.
Hence, consider a more complicated learning problem.

Suppose only have the observations $\theta(u_0), \theta(u_1), \ldots, \theta(u_m)$ of the true system states $u$ in an unknown dynamical system $(U,T)$ are available and, we wish to predict the values $\theta(u_{m+1})$, $\theta(u_{m+2})$ and so forth.

We start by establishing the method of Takens delay embedding. 
In this method, one considers a discrete-time dynamical system defined on a `nice' space 
(a smooth manifold~\footnote{A manifold is understood to be a special topological space that is locally homeomorphic at each point to an open set of $\mathbb{R}^d$})
that can be obtained as a time-$K$ map of a continuous-time dynamical system, concepts which are formalised below.
We make observations from such a system, i.e., we observe the evolution of $\theta(w_0)$ (where $w_0$ represents the initial state) by examining  a finite set of values of $u_n :=  \theta(w_n) = \theta(Tw_{n-1})$. The sequence $\{u_n\}$ represents a scalar time-series. Intuitively $\theta$ represents a probe inserted into the larger system which is itself only measuring/extracting a small part of the greater system state at time $t=n$. 
Consider, for example, a thermometer erected to measure the ambient temperature in a local village. This measurement function- the thermometer- captures only a single aspect- the temperature- of a much grander dynamical system encompassing the current weather of the surrounding area. Even more than that- this probe is measuring a miniscule part of the global weather system.

@@

As already mentioned, the actual state $u$ of a system is nearly never fully known. The most one can do is to place probes into a system to obtain partial information through the measurements taken. Moreover, the process of taking a measurement itself introduces two additional aspects which complicate the problem. 
\vspace{-6mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item A series of measurements over a specific time interval is inherently a discretisation procedure of the underlying continuous-time dynamical system. (Hence why we restricted our attention to discrete-time systems in the preceding section)
  \item A probe will never be entirely accurate; the act of measurement introduces a certain measure of numerical noise/inaccuracy.
\end{enumerate}

Takens embedding theorem does not take into account the noise, and to describe it we construct a multidimensional observable using the method of stacking previous observations.
The concept of an observable is understood in the sense of an observable as originally introduced by~\cite{takens1981detecting, genericObservableAeyels} to refer to a probe or measurement function inserted into the system. 
When speaking of stacking, we mean that a delay-coordinate map is defined by
$\Phi_{k,\theta}(w) := (\theta(T^{-k}w)\ldots,\theta(T^{-1}w),\theta(w))$.  
In essence, Takens theorem refers to the fact that for a sufficiently large $k$, we can define a dynamical system on the space $\mathbb{R}^{k+1}$ with states evolving as $\Phi_{k,\theta}(w)$, $\Phi_{k,\theta}(Tw)$, $\Phi_{k,\theta}(T^2w)$, etc. This dynamical system is topologically conjugate to the unknown underlying system $(W,T)$. We recall the Takens delay embedding theorem next.

@@

We define first the concepts of homeomorphism and embedding.
\begin{Definition} \rm
  [\bf {Homeomorphism}]\label{Dfn_homeo}\rm
  A homeomorphism is a function $f:Z\rightarrow Y$ between two topological spaces $Z$ and $Y$ that is continuous, bijective and has a continuous inverse. 
\end{Definition}

@@

@@

Takens Theorem states a result establishing a relationship between the observed and underlying dynamical systems by showing that the concatenation of a sufficiently large number of previous observations into a vector will, under certain conditions, generate a map between the vectors from the respective systems.  We formulate the theorem from~\cite{takens1981detecting}.

\begin{Theorem} 
	[\bf Takens Embedding Theorem (taken from~\cite{takens1981detecting})]\label{Thm_Takens}
         Let $W$ be a compact manifold of dimension $m$, and $d\ge m$ so that $2d$ is an integer. It is a 
            generic property for the pair $(T, \theta)$,  where $T:W \to W$ is
            a smooth diffeomorphism, and $\theta:W \to \mathbb{R}$ a smooth function, the map $\Phi_{2d,\theta}:W \to \mathbb{R}^{2d+1}$ defined on $W$ by 
            $\Phi_{2d,\theta}(w) := (\theta(T^{-2d}w)\ldots,\theta(T^{-1}w),\theta(w))$
            is a diffeomorphic embedding. Consequently, there exists a map $F_\theta: \Phi_{2d,\theta}(W) \to \Phi_{2d,\theta}(W)$ defined by \[F_\theta: (\theta(T^{-2d}w),\ldots,\theta(T^{-1}w),\theta(w)) \mapsto 
            (\theta(T^{-2d+1}w),\ldots,\theta(w),\theta(Tw)) \]
           so that $(W,T)$ is topologically conjugate to 
            $(\Phi_{2d,\theta}(W), F_\theta)$.    
\end{Theorem}

By generic, we mean a residual set on a certain topology on appropriate spaces of functions (that we do not describe here) \ednote{B: Where can I find a citation for this?}. 
In our scenario -- and this remains important throughout -- the input space $U$ is considered the attractor.~\label{attractor_U}

@@

\begin{figure}[ht]
  \includegraphics[width=0.7\linewidth]{Graphs/_takensmap.eps}
  \centering
  \captionof{figure}{Schematic of Takens Embedding Theorem's usage and impact of iterates slipping from the attractor when the approximation $\tilde{F}_\theta$ of $F_\theta$ is learnt.}
 \label{fig:takensmap}
\end{figure}

@@

\section{Practical Limitations to Takens' Theorem}\label{sect_takenslimits}
The theorem above is a powerful result, and it provides compelling reason to believe that one could conceivably accurately reconstruct a system conjugate to the underlying system. 
Nonetheless, it does present some severe practical limitations.
\vspace{-5mm}
\begin{enumerate}
\item Even supposing that we could indeed find $F_\theta$, our \emph{approximation} of  $F_\theta$ is a map from a larger set $\mathbb{R}^{2d+1}$ containing the embedded attractor. There are, however, no theoretical guarantees that $F_\theta$ will retain  $V:=\Phi_{2d,\theta}(W)$  as an attractor although $W$ could itself be an attractor.
\item Takens theorem is stated only for noiseless observations. Due to noise $\epsilon_n$, the delay vector $\Phi_{2d,\theta}(w_n) + \epsilon_n$  may lie outside $V$. Furthermore, due to the chaotic nature of the underlying system (i.e. the fact that it has SIDC), the evolution of $\Phi_{2d,\theta}(w_n) + \epsilon_n$ under the map $F_\theta$ could move out of $V$ completely. This problem can be overcome  by using a driven dynamical system with specific properties, and we discuss this in Chapter~\ref{ch3}. 
\end{enumerate}

@@

Though not uniquely related to the Takens embedding theorem, we opt to make mention of another requirement to our work here. 
The methods described in this project are applicable for data originating from a surjective map. 
This follows from the fact that when $T$ maps the space $U$ into a proper subset $B$ of $U$, there exists some point in $U - A$ that does not have a preimage such that $T$ cannot be not surjective. 
An example of this would be the case when a system displays energy-loss. Consider, for instance, consider the dying oscillations for a damped pendulum (see Figure~\ref{fig:damped_pendulum}). The double pendulum is discussed in greater depth  is provided in Chapter~\ref{ch5}.
The requirement is not overly restrictive as map that is topologically transitive on a subspace must also be surjective on it; if it were to be not surjective there would be an open set $V$ that does not intersect with any future iterates of an open set  $U$ disjoint from it.
